---
title: "STAT 425 Final Project"
author: "Jack Liu, Reed Haas"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

::: {align="center"}
# **A study of students academic performance Factor**
:::

#### Group Member Contribution:

-   Jack (EDA and modeling)

-   Reed (model diagnostics and model comparisons)

-   Jack & Reed (conclusion and discussion)

## Abstract

## Exploratory Data Analysis

### data preparation

```{r}
df = read.csv("/Users/jacksandroids/Documents/UIUC/Spring 2025/stat 425 R Code/final project/StudentsHabits.csv")
View(df)
print("Drop column 'X'")
df <- df[, -1]
```

```{r}
cat("The shape/dimension of the student performance dataset is", dim(df))

cat("\n", "the sample size is", nrow(df), " the number of predictors:", ncol(df) - 1)

```

```{r}
print_model_fit <- function(model) {
  model_summary <- summary(model)
  model_name <- deparse(substitute(model))
  cat("model fit informations for", model_name, "\n")
  cat("Residual Standard Error:", model_summary$sigma, "\n")
  cat("R-squared:", model_summary$r.squared, "\n")
  cat("Adjusted R-squared:", model_summary$adj.r.squared, "\n")
  cat("F-statistic:", model_summary$fstatistic[1], "on", 
      model_summary$fstatistic[2], "and", model_summary$fstatistic[3], 
      "DF, p-value:", pf(model_summary$fstatistic[1], 
                         model_summary$fstatistic[2], 
                         model_summary$fstatistic[3], 
                         lower.tail = FALSE), "\n")
  cat("\n")
}
```

```{r}
model_1 = lm(formula = exam_score ~ . , data = df)
summary(model_1)

model_intercept = lm(formula = exam_score ~ 1 , data = df)
```

```{r}
library(GGally)
library(ggplot2)
library(faraway)
library(regclass)
library(ggplot2)
library(MASS)
library(lmtest)
library(boot)
library(MASS)
library(leaps)
library(splines)
```

### numerical

Explore categories, variable types, numerical summaries with demonstrations on overall distributions; additionally, we also explore factor level composition in all categorical variables

```{r}
cat("the column names are",colnames(df) , "\n", "info of each variable",  str(df))
```

```{r}
summary(df)
```

```{r}
print("categorical variables analysis" )
print("extract all categorical variables")
categorical_vars <- c()
categorical_vars <- names(df)[sapply(df, function(x) is.character(x) || is.factor(x))]
categorical_vars
for (var in categorical_vars) {
  #cat("---", var , "---")
  print(table(df[var]))
  cat("\n")
}

```

**observations:**

#### 

## graphical:

```{r}
numeric_var <- colnames(df)[sapply(df, is.numeric) & colnames(df) != "exam_score"]
par(mfrow = c(2, 4)) 
for (var in numeric_var) {

  
    main_title <- paste("Predicting with", var)
    x_label <- var

    plot(x = df[[var]] , y = df$exam_score, main = main_title, xlab = x_label, ylab = "exam_score", pch = 19, col = "blue")
    lm_model <- lm(as.formula(paste("exam_score ~", var)), data =df)
    abline(lm_model, col = "red", lwd = 4) 
  
}
```

```{r}
# Get the list of categorical variables (excluding "exam_score")
cat_vars <- colnames(df)[sapply(df, is.character) & colnames(df) != "exam_score"]

# Set up plotting area: adjust rows (e.g., 2) and columns (e.g., 3) depending on how many plots
par(mfrow = c(2, 3))  
# Loop through categorical variables
for (var in cat_vars) {
  main_title <- paste( var)
  x_label <- var

  boxplot(df$exam_score ~ df[[var]],
          drop = FALSE,
          col = "steelblue",
          names = levels(factor(df[[var]])),
          xlab = x_label,
          ylab = "exam_score",
          main = main_title)
}
```

correlation graphs, non linear pattern, cross examined between categorical and numerical variables

```{r}
# Correlation Graph 

#ggpairs(df)

print("Correlation Assessment")

# Insepct correlation in numerical prediction variables

df_numeric <- df[sapply(df, is.numeric)]
df_cate <- df[sapply(df, is.character)]
corr_matrix <- cor(df_numeric[, -9])
corr_matrix
print("overall very small correlation value between all numerical predictor variables, chances of multicollinearity in numerical variables in small")

```

```{r}
# df_cate should be a character vector of column names
df_cate <- c("gender", "part_time_job", "diet_quality", 
             "parental_education_level", "internet_quality", 
             "extracurricular_participation")

# Loop through all unique pairs
for (i in 1:(length(df_cate) - 1)) {
  for (j in (i + 1):length(df_cate)) {
    var1 <- df_cate[[i]]  # use double brackets to get string
    var2 <- df_cate[[j]]
    
    tbl <- table(df[[var1]], df[[var2]])
    test <- chisq.test(tbl)
    
    cat( var1, "vs", var2, "      ", round(test$p.value, 5), "\n")
  }
}

print("We use Chi-Square test checks whether two categorical variables are independent (null hypothesis) or associated  (alternative hypothesis) at significance level of 0.05")
print("we fail to reject the Chi-squared test's null hypothesis and conclude that there is no strong assciation between all categorical variables")


```

```{r}



#vif(model_1)
round(VIF(model_1), dig=5)
```

Conclusion: all variables in the student performance dataset have low correlations, so we can carry out the modeling stage safely.

## Model Building

### 1. Linear model with all variables, without interactions (model 1)

```{r}

anova(model_1, model_intercept)
```

*Context:*

For the initial linear modeling of "exam_score", I fit all column variables, including various factor levels in categorical columns, to predict the linear regression of "exam_score".

*Finding:* this linear model does explain significant variation in response variable ("exam_score") when it compares against an intercept-only linear model, which implies certain predictors in the model is significant in predicting "exam_score" variable.

assumptions needed to review: linearity, independence of error, homosedasticity, normality of errors, no multicollinearity (checked)

### 2. Outliers Diagnosis

```{r}
lev <- influence(model_1)$hat
p <- ncol(df); n <- nrow(df)
lev[lev > (2 * p / n)]
#high leverage points
# halfnorm(lev, length(lev[lev > (2 * p / n)]), labs=row.names(df), ylab="Leverage")
print("observed students that have high-leverage data")
df[lev > (2*p/n), ]

```

```{r}
print("check for outliers")
print("First use studenized residuals")
rstud <- rstudent(model_1)
cutoff <- abs(qt(0.05/(2*dim(df)[1]), model_1$df.residual-1))
paste("the Bonferroni corrected outlier cutoff value is", cutoff)
sort(abs(rstud), decreasing=TRUE)[1:10]
# utlier_points <- which(sort(abs(rstud), decreasing=TRUE)[1:10] > cutoff) 
paste("outlier points in Bonferroni correction are observed 226th, 472th, and 431th observations; Totally 3 high outlier points in 65 high leverage points")

paste("the cutoff without Without Bonferroni correction is",abs(qt(0.05/2, model_1$df.residual-1)))

paste("Cook distance graph")
cook = cooks.distance(model_1)
halfnorm(cook, labs = row.names(df), ylab = "Cook's distance")
which.max(cook)
```

```{r}
df_new <- df[-c(226, 431, 472), ]

df_no_high_leverage <- df[lev <= (2 * p / n), ]
model_no_lev <- lm(exam_score ~ ., data = df_no_high_leverage)
model_new <- lm(exam_score ~ ., data = df_new)
print_model_fit(model_no_lev)
print_model_fit(model_new)
print_model_fit(model_1)
paste("drop outlier points (226th, 431th, and 472th observations) only ")
```

### 3. Diagnose Linearity & Homoscedasticity (equal variance)

```{r}
par(mfrow=c(2,2))
plot(model_new)
```

```{r}
plot(fitted(model_new), residuals(model_new),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values",
     pch = 19, col = "steelblue")
abline(h = 0, lwd = 2, col = "red")


ggplot(data = data.frame(fitted = fitted(model_new),
                         residuals = residuals(model_new)),
       aes(x = fitted, y = residuals)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal(base_size = 14) + geom_smooth()


```

### 4. Box-Cox Transformation Diagnosis

**log transformation, delete later**

```{r}

df_log <- df_new
numeric_var <- sapply(df_log, is.numeric) & names(df_log) != "exam_score"
df_log[numeric_var] <- log(df_log[numeric_var] + 1)
model_log <- lm(exam_score ~ ., data = df_log)
print_model_fit(model_log)
print_model_fit(model_new)
```

```{r}
paste("ggplot clearly demonsrtates non-linearity and unequal variance due curved pattern, high residual deviations")
paste("remedy: boxcox transformation model")


bc <- boxcox(model_new, plotit=TRUE, lambda = seq(1,5,0.05))
boxcox(model_new, plotit=TRUE, lambda = seq(1,5,0.05))
lambda_optimal <- bc$x[which.max(bc$y)]
cat("Optimal lambda value is", lambda_optimal)

```

**apply lambda transformation using this equation**

$$
g(y) = \frac{y^{\lambda} - 1}{\lambda}
$$

```{r}
#df_new$exam_score <- (((df_new$exam_score)**(lambda_optimal) - 1) / lambda_optimal)
model_bc <- lm(( ((exam_score)^(lambda_optimal) - 1) / lambda_optimal) ~ . , data =  df_new)
print_model_fit(model_bc)
cat("\n")
print_model_fit(model_new)

print("Box-Cox transformation improves model fit")
```

```{r}
par(mfrow=c(2,2))
plot(model_bc)
```

```{r}
# Load spline function


# Fit a model using natural splines for top 5 predictors
model_spline <- lm(
  exam_score ~ 
    ns(study_hours_per_day, df = 4) + 
    ns(mental_health_rating, df = 4) +
    ns(sleep_hours, df = 4) +
    ns(social_media_hours, df = 4) +
    ns(exercise_frequency, df = 4) +
    gender + part_time_job + 
    attendance_percentage + diet_quality + 
    parental_education_level + internet_quality + 
    extracurricular_participation + age,
  data = df
)

summary(model_spline)
```

### 

### 5. Independence of Error diagnosis

```{r}

dwtest(model_bc, alternative = "two.sided")
paste("there is no evidence of autocorrelation in the residuals and the errors appear to be independent")
```

### 6. Normality diagnosis

```{r}
shapiro.test(residuals(model_bc))
shapiro.test(residuals(model_new))

plot(model_bc, which = 2)
paste("Remedy: use GLM  to diagnose non-normality issue")

model_glm <- glm(exam_score ~ ., data = df_new)
summary(model_glm)
# print_model_fit(model_glm)

```

```{r}
cat("remove outlier points 568th, 779th, and 23rd in QQ-residuals graph")
df_new <- df_new[-c(568, 779, 23), ]
cat("Remedy: bootstrap")


boot_fn <- function(data, indices) {
  d <- data[indices, ]
  coef(lm(exam_score ~ ., data = d))
}
boot(df_new, boot_fn, R = 1000)
```

```{r}
model_ols <- lm(exam_score ~ ., data = df_new)

model_robust <- rlm(exam_score ~ ., data = df_new)
ols_coef <- coef(model_ols)
robust_coef <- coef(model_robust)
diff <- abs(ols_coef - robust_coef)
print(sort(diff, decreasing = TRUE))
paste("	•	Large differences (e.g., in Intercept, parental_education_levelMaster, study_hours_per_day) suggest that outliers or non-normality are strongly influencing these predictors, we will further investigate those predictors using Variable selection")
```

```{r}
shapiro.test(resid(model_ols))
shapiro.test(resid(model_robust))
```

### 

### 

### 7. Variable Selection

```{r}
cat("use leap & bound algorithm to determine best model size")

leap = regsubsets(exam_score ~., data = df_new)
length(leap)
#rs = summary(leap)
#rs$which

selected_model <- summary(leap)$which 
model_size <- rowSums(selected_model)

## [1] "Displays selected variables in model size 2 to 8 A formatted table"
selected_model[model_size %in% 2:8,]
```

### 7. ANCOVA Model with Interactions (Model Comparison with Model 4 Included)

### 8. Shrinkage Methods (Ridge and Lasso Regression)

### 

## Discussion of Results and Conclusions

Based on a series of model building, diagnostics and comparisons, we find that **the following variables are important in determining the performance of a student, measured by the student's final exam score**
