---
title: "STAT 425 Final Project"
author: "Jack Liu, Reed Haas"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {align="center"}
# **A study of students academic performance Factor**
:::

#### Group Member Contribution:

-   Jack (EDA and modeling)

-   Reed (model diagnostics and model comparisons)

-   Jack & Reed (conclusion and discussion)

## Abstract

## Exploratory Data Analysis

### data preparation

```{r}
df = read.csv("/Users/jacksandroids/Documents/UIUC/Spring 2025/stat 425 R Code/final project/StudentsHabits.csv")
View(df)
print("Drop column 'X'")
df <- df[, -1]
```

```{r}
cat("The shape/dimension of the student performance dataset is", dim(df))

cat("\n", "the sample size is", nrow(df), " the number of predictors:", ncol(df) - 1)

```

### numerical

Explore categories, variable types, numerical summaries with demonstrations on overall distributions; additionally, we also explore factor level composition in all categorical variables

```{r}
cat("the column names are",colnames(df) , "\n", "info of each variable",  str(df))
```

```{r}
summary(df)
```

```{r}
print("categorical variables analysis" )
print("extract all categorical variables")
categorical_vars <- c()
categorical_vars <- names(df)[sapply(df, function(x) is.character(x) || is.factor(x))]
categorical_vars
for (var in categorical_vars) {
  cat("\n---",var , "---\n")
  print(table(df[var]))
  cat("number of observations is", nrow(df[var]), "\n")
  
}

```

**observations:**

#### 

## graphical:

```{r}
for (var in colnames(df)) {
  if (is.numeric(df[[var]])  &&  var != "exam_score") {
  
    main_title <- paste("Predicting exam scores with", var)
    x_label <- var

    plot(x = df[[var]] , y = df$exam_score, main = main_title, xlab = x_label, ylab = "exam_score", pch = 19, col = "blue")
    lm_model <- lm(as.formula(paste("exam_score ~", var)), data =df)
    abline(lm_model, col = "red", lwd = 4) 
  }
}
```

```{r}
for (var in colnames(df)) {
  if (is.character(df[[var]]) &&  var != "exam_score") {
  
    main_title <- paste("Boxplot distribution of students' final score on ", var, "factor levels")
    x_label <- var

    boxplot(df$exam_score ~ df[[var]],drop = FALSE, col = "steelblue",names = levels(factor(df[[var]])),  xlab = x_label, ylab = "exam_score")
    
   
  }
}
```

correlation graphs, cross examined between categorical and numerical variables

## Model Building

### 1. Linear model with all variables, without interactions (model 1)

```{r}
model_1 = lm(formula = exam_score ~ . , data = df)
summary(model_1)

model_intercept = lm(formula = exam_score ~ 1 , data = df)
anova(model_1, model_intercept)
```

*Context:*

For the initial linear modeling of "exam_score", I fit all column variables, including various factor levels in categorical columns, to predict the linear regression of "exam_score".

*Finding:* this linear model does explain significant variation in response variable ("exam_score") when it compares against an intercept-only linear model, which implies certain predictors in the model is significant in predicting "exam_score" variable.

### 2. Diagnostics for the basic linear model

### 3. Refit a linear model, (model 2)

### 4. Diagnostics for the refitted linear model (model 2)

### 5. Build a model with removed outliers (model 3)

### 6. Variable Selection (Model Comparison with Model 3 Included)

### 7. ANCOVA Model with Interactions (Model Comparison with Model 4 Included)

### 8. Shrinkage Methods (Ridge and Lasso Regression)

### 

## Discussion of Results and Conclusions

Based on a series of model building, diagnostics and comparisons, we find that **the following variables are important in determining the performance of a student, measured by the student's final exam score**
