---
title: "STAT 425 Final Project"
author: "Jack Liu, Reed Haas"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {align="center"}
# **A study of students academic performance Factor**
:::

#### Group Member Contribution:

-   Jack (EDA and modeling)

-   Reed (model diagnostics and model comparisons)

-   Jack & Reed (conclusion and reed preparation)

## Abstract

## Exploratory Data Analysis

#### data preparation

```{r}
df = read.csv("/Users/jacksandroids/Documents/UIUC/Spring 2025/stat 425 R Code/final project/StudentsHabits.csv")
View(df)
```

```{r}
cat("The shape/dimension of the student performance dataset is", dim(df))

cat("\n", "the sample size is", nrow(df), " the number of predictors:", ncol(df) - 1)
```

#### numerical 

#### 

#### graphical

## Model Building

### 1. Linear model with all variables, without interactions (model 1) 

### 2. Diagnostics for the basic linear model 

### 3. Refit a linear model, (model 2)

### 4. Diagnostics for the refitted linear model (model 2)

### 5. Build a model with removed outliers (model 3) 

### 6.  Variable Selection (Model Comparison with Model 3 Included) 

### 7. ANCOVA Model with Interactions (Model Comparison with Model 4 Included) 

### 8. Shrinkage Methods (Ridge and Lasso Regression) 

### 

## Discussion of Results and Conclusions

Based on a series of model building, diagnostics and comparisons, we find that **the following variables are important in determining the performance of a student, measured by the student's final exam score**
